---
title: 'Plots for Michal''s manuscript: Proteomics results'
author: "Friederike Dündar"
date: "8/29/2019; updated `r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r cache=FALSE}
library(data.table)
library(magrittr)
library(ggplot2); theme_set( theme_bw(base_size = 16) )
library(clusterProfiler)
library(pheatmap); library(RColorBrewer)
```

Make sure to download the file `pxClusters_logFC-DNvsYA_perplex40_dataTable_with_kNNClusters.rds` from here: <https://wcm.box.com/s/46ixb9uraxsnkug6q68hok61kwiohk4q>

```{r px_load_clustering_and_tsne_results, cache=FALSE}
t_px <- readRDS(file = "pxClusters_logFC-DNvsYA_perplex40_dataTable_with_kNNClusters.rds")
i <- 20
```

Based on the similarities , we're going to combine the following px clusters:

```{r combining_clusters, cache=FALSE}
t_px[, combClust := ifelse(cluster_k20 == 10 | cluster_k20 == 11, "A",
                    ifelse(cluster_k20 == 3, "B",
                    ifelse(cluster_k20 == 4 | cluster_k20 == 8 | cluster_k20 == 6, "C", 
                    ifelse(cluster_k20 == 1, "D",
                    ifelse(cluster_k20 == 2, "E",
                    ifelse(cluster_k20 == 9 | cluster_k20 == 7 | cluster_k20 == 5,"F", 
                           NA))))))]
```


```{r vioplot_combinedClusters_tx, fig.width = 15, fig.height = 7}
## define colors
my_cols_combClust <- c("dodgerblue1","blue","navyblue","darkorchid1","deeppink1","firebrick2","firebrick4")

ggplot(t_px, aes(x = day, y = logFC)) + 
  geom_violin() + 
  ggbeeswarm::geom_quasirandom(aes(color = combClust), groupOnX = TRUE) + 
  facet_grid(~combClust) +
  ggtitle("Expression changes per day (Proteomics)") +
  scale_color_manual(name = "Cluster", values = my_cols_combClust) +
   guides(colour = guide_legend(override.aes = list(alpha = 1, size = 5, shape = 15)))
```



## ORA of HALLMARK gene sets

We used `compareCluster()` and `enricher()` of the `clusterProfiler` R package to determine which gene sets of the MSigDB Hallmark data set were overrepresented.
The enrichment score is based on a hypergeometric test.
The comparison function of clusterProfiler is designed as a general framework for comparing any kind of biological and biomedical ontologies and automatically calculating functional categories that are enriched in each gene cluster (p-values < 0.01 and FDR < 0.05) [(Yu et al., 2012)](https://dx.doi.org/10.1089/omi.2011.0118)

More enrichment analyses: `2019-04_GOterms_of_Clusteres.Rmd`

```{r defining_universe_prots}
load("prots_norm.rda") #from data("prots_norm", package = "BlenisPxData")
eg.p <-  clusterProfiler::bitr(unique(prots_norm$uniprot), fromType="UNIPROT", 
                               toType="ENTREZID", 
                               OrgDb="org.Hs.eg.db") %>% as.data.table
setnames(eg.p, names(eg.p), c("uniprot", "entrez"))
```

```{r}
clstcomp2.px <- eg.p[t_px, on = "uniprot"] %>% 
  .[, c("gene_symbol","entrez","combClust","day","logFC"), with=FALSE] %>% 
  .[!is.na(entrez)] %>% unique
clstcomp.px.list22 <- lapply(sort(unique(clstcomp2.px$combClust)), 
  function(x) clstcomp2.px[combClust == x]$entrez )
names(clstcomp.px.list22) <- sort(unique(clstcomp2.px$combClust))
```

```{r prep_ORA, cache=TRUE}
hmgs <- clusterProfiler::read.gmt(  "h.all.v6.2.entrez.gmt")
```

### Comparing the (summarized) clusters

```{r run_compCluster_on_combinedClusters_Hallmark, cache=TRUE}
hmgs <- clusterProfiler::read.gmt(  "h.all.v6.2.entrez.gmt")
ck2.Hallmark <- compareCluster(clstcomp.px.list22, fun = "enricher", TERM2GENE = hmgs,  universe = unique(eg.p$entrez))
```

```{r dotplot_overrepHallmark_combClust, fig.width = 8, fig.height= 7}
dotplot(ck2.Hallmark) + 
  ggtitle("Overrepresented HALLMARK gene sets for manually combined clusters") +
  scale_color_gradientn(colours = rev(c("cornsilk1", "firebrick3", "firebrick4")),
                        limits = c(0,0.05)) +
  theme(legend.position = "bottom")
```
