---
title: "Identifying differentially expressed proteins"
author: "Friederike Duendar"
editor: "Michal Nagiec"
date: "3/28/2019"
output:
    html_document:
      toc: true
      toc_depth: 4
      toc_float: true
      code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=FALSE)
```

```{r cache=FALSE}
options(stringsAsFactors = FALSE)
library(magrittr)
library(data.table)
library(readxl)
library(ggplot2)
```

>These data sets focus on one time point where we have three replicates built into the experiment and usually comparing three experimental conditions.
> A data set comparing **MCF10A cells (control, ERK-DN expressing and TGF-beta treated)**. 
>In the sheet you will see that the raw data is in columns E-M and further normalized in the following columns.
>E-G are EV replicates, H-J are DN replicates and K-M are TGF replicates.

# Reading in the data

```{r}
## preparing the reading of the xls file
filename <-  "DNvTGF_tPx_PC2.xlsx"
xls_path <- paste0(filename)
```

```{r}
## reading in the xls file
prots.in <- readxl::read_excel(xls_path, sheet = 1, col_names = TRUE, skip = 0) %>% data.table
names(prots.in)
```

```{r}
# we need to remove empty spaces etc. to make the column names compatible with other R functions
setnames(prots.in, names(prots.in), make.names(names(prots.in)))

# remove the second Gene ID column at the end of the table
prots.in <- prots.in[, -"Gene__1"  , with=FALSE]
```


Turning the wide table into a long format:

```{r}
prots <- melt.data.table(prots.in, id.vars = 1:4) # id.vars should specify the columns with details that are common to a single protein, i.e. Protein IDs, Number of peptides
prots
```

The `variable` column now contains the headers of the columns with values.

```{r}
unique(prots$variable)
# we don't want factors for the experiment names yet
prots$variable <- as.character(prots$variable)
# only keep the __raw__ values for now -- these are stored in the rows corresponding to variables that start with "X"
prots <- prots[variable %in% grep("^X", unique(prots$variable), value = TRUE)]
# change the gene column header
setnames(prots, "Gene", "gene_symbol")
```

## Cleaning the data

```{r}
# remove lines without protein IDs
prots <- prots[!is.na(Protein.Id)]
```

### Checking and removing contaminants

```{r}
prots[grepl("ACHLY", Protein.Id)]
prots <- prots[!grepl("ACHLY", Protein.Id)]
```

```{r}
prots[!grep("^sp", Protein.Id)]
prots <- prots[grep("^sp", Protein.Id)]
```

```{r}
prots[!grepl("HUMAN", Protein.Id)]
prots <- prots[grepl("HUMAN",Protein.Id)]
```

## Adding sample information

```{r}
# defining samples based on the printed names
sample_names <- c("EV1","EV2","EV3","DN1","DN2","DN3","TGFB1","TGFB2","TGFB3")
col_names <- unique(prots$variable) 
samples <- data.table(sample = sample_names,
                      condition = gsub("[0-9]","", sample_names),
                      variable = col_names)
samples
```

Joining the sample info table with the raw values

```{r}
prots <- samples[prots, on = "variable"]
```

```{r eval=FALSE}
#saveRDS(prots, file ="protein_quant_ERKvsTGB_2019-03-27_exp2.RDS")
```


# Normalization

Jan Krumsiek first applies a protein-focussed median polish, followed by the calculation of a dilution factor where
`d = median(value/median(protein))`.

```
# get the median value for every protein [molecules are stored in columns here]
  ref <- apply(X, 1, function(x) median(x,na.rm=T))

# get dilution factors (per sample)
  d <- apply(X, 2,  function(s) median(as.numeric(s/ref),na.rm=T))

# apply to each sample (for each row=sample, divide values by median dilution factor)
  Y = sapply(1:dim(X)[2], function(i)unlist(X[,i]/d[i]))
```

```{r}
make_KrumScript_matrix <- function(in.dt, which_ID = "metabolite", which_value = "value"){

  if(which_ID == which_value){stop("which_ID and which_value must specify different columns.")}

  ABCutilities::check_columns(c("sample",which_ID, which_value), in.dt, "in.dt", "make_KrumScript_matrix")

  km <- data.table::dcast(in.dt[, c("sample",which_ID, which_value), with = FALSE],
                          as.formula(paste0("sample~", which_ID)),
                          value.var = which_value) %>% data.frame()
  rownames(km) <- make.names(km$sample)
  colnames(km) <- make.names(names(km))
  km$sample <- NULL
  return(km)
}
```

```{r}
medianNorm_metabolites <- function(X, batches = NULL){

  if (is.null(batches)){
    message("No batches given, therefore we will use the median per molecule across all samples.")
    #batches <- ""
    X <- apply(X, 2, function(c){ c / median(c,na.rm=T)})
  }else{
    if(length(batches) != dim(X)[1]){
      stop("The batches vector should indicate the batch for every sample (row) of X.")
    }
   for (i in 1:length(levels(batches))) {
    batch <- levels(batches)[i]
    # median normalize every metabolite within the same batch
    Xnorm = apply(X[batches==batch,], 2, function(c){ c / median(c,na.rm=T)})
    X[batches==batch,] = Xnorm
   }
  }

  return(X)
}

```

```{r}
quotNorm = function(X, vars=1:dim(X)[2], NAerror=F, refsamples=NA) {
  # x:          data frame to be normalized
  # vars:       index vector of variables o be used, default: all
  # NAerrors:   throw error for NA's or just ignore?
  # refsamples: indices of samples to calculate reference sample on (e.g. only on control samples)
  
  # crash if there are any negative values
  if (any(X[!is.na(X)]<0)) stop("Matrix contains negative values. Did you input logged data?")
  
  # check if there are any NAs
  if (sum(is.na(X[,vars]))>0) {
    # throw warning or error?
    if (NAerror) {
      stop('Data matrix contains NAs')
    } else {
      warning('Data matrix contains NAs')
    }
  }
  
  # if reference samples not given -> all samples
  if (any(is.na(refsamples))) refsamples <- 1:nrow(X)
  
  # median reference sample
  ref = apply(X[refsamples,vars],2,function(x)median(x,na.rm=T))
  # get dilution factors
  d = apply(X[,vars],1,  function(s) median(as.numeric(s/ref),na.rm=T))
  # apply to each sample  (for each row=sample, divide values by median dilution factor)
  Y = t(sapply(1:dim(X)[1], function(i)unlist(X[i,]/d[i])))
  
  #Y = t(apply(X,1,  function(s) s /  d) )
  rownames(Y) = rownames(X)
  
  # return
  list(X=Y,dilution=d)
}

```

```{r jan_norm}
norm_matrix_list <- list()
## need to transform the long data.table into a matrix where rows correspond to samples
norm_matrix_list$ori.value <- make_KrumScript_matrix(prots,
                                                     which_ID = "Uniprot",
                                                     which_value = "value") 
## normalize protein range per MS run (=batch)
norm_matrix_list$median_norm <- medianNorm_metabolites(norm_matrix_list$ori.value, 
                                                                  batches = NULL # we don't have different MS runs here!
                                                                  )
## adjust sample ranges
norm_matrix_list$quot_norm <-quotNorm(norm_matrix_list$median_norm ) %>% .$X

# turn into data.tables
combi.dt <- lapply(seq_along(norm_matrix_list), function(i){
  dt <- reshape2::melt(as.matrix(norm_matrix_list[[i]]))
  dt <- as.data.table(dt)
  setnames(dt, c("Var1","Var2"), c("sample", "uniprot"))
  dt$type <- names(norm_matrix_list[i])
  return(dt)
  }
  ) %>% rbindlist %>% dcast(., uniprot + sample ~ type, value.var = "value")

combi.dt <- combi.dt[samples, on = "sample"][!is.na(ori.value)]

```

```{r define_colors}
condition_cols <- c("turquoise","dodgerblue1","navyblue")
replicate_cols <- c("violetred1","violet", "purple") 
```

```{r eval=FALSE}
saveRDS(combi.dt,file ="protein_quant_normalized_ERKvsTGB_2019-03-27_exp2)
```

# Differential protein analysis

```{r}
library(limma)
```

For limma, we will need a matrix in wide format again.

```{r}
prot.mat <- dcast(combi.dt,  uniprot ~ sample, value.var = "quot_norm") %>% as.data.frame
row.names(prot.mat) <- prot.mat$uniprot
prot.mat$uniprot <- NULL
prot.mat <- as.matrix(prot.mat)
head(prot.mat)
```

### Design matrix

Then, we need to define the conditions that we're going to compare.

```{r}
cnames <- colnames(prot.mat)
grps <- factor(gsub('[1-3]$', '', cnames))
grps
#[1] DN   DN   DN   EV   EV   EV   TGFB TGFB TGFB
#Levels: DN EV TGFB
#sub.matDN <- prot.mat[, grep("DN|EV", colnames(prot.mat), value = TRUE), drop = FALSE]
#sub.matTGFB <- prot.mat[, grep("TGFB|EV", colnames(prot.mat), value = TRUE), drop = FALSE]
#cnames <- colnames(sub.mat)
```

The design.matrix is an integral component of limma; if you want to understand the
rationale behind this, I recommend the chapters on Matrix Algebra and Linear Models
from the book [Data Analysis for the Life Sciences by Rafa Irizarry and Michael Love](https://leanpub.com/dataanalysisforthelifesciences).

The design matrix is meant to capture which sample belongs to which condition.

```{r}
design.matrix <- model.matrix(~ 0 + grps)
colnames(design.matrix) <- as.character(levels(grps))
row.names(design.matrix) <- cnames
design.matrix
```

### Contrast matrix

In addition, we need the `contrast.matrix`
```{r}
#limma::makeContrasts( (DN-EV)-(TGFB-EV) , levels = design.matrix)
contrast.matrix <- limma::makeContrasts(DN-TGFB, DN-EV, TGFB-EV, levels = design.matrix)
contrast.matrix
```


### Perform the DE analysis

Now, we have everything we need:

* a matrix of values where rows are proteins and columns are samples (`prot.mat`)
* a design matrix where every row corresponds to the samples of `prot.mat`
* a contrast matrix to specify how the log2FC values are supposed to be calculated

```{r}
pm <- as.matrix(prot.mat)
## replace NA values with a small pseudo-value
pm[is.na(pm)] <- log2(2^(min(pm, na.rm = TRUE))/2)

## fit the linear model to the observed values
fit <- limma::lmFit(pm, design.matrix)
## estimate the coefficients of the model, i.e. generating logFC values
fit2 <- limma::contrasts.fit(fit, contrast.matrix)
## perform the statistical test of whether the logFC is "far away" from zero
## for a given protein
fit2 <- limma::eBayes(fit2)
## extract the results: return all proteins that pass the q-value cut-off
```


```{r prep_extraction_of_DP}
# make a data.table with additional identifiers that will be added to the proteins
# of interest later on
prot_anno <- prots[, c("Protein.Id","Uniprot","gene_symbol"), with=FALSE] %>% unique
prot_anno$protein <- gsub(".+\\|([A-Z0-9]+)_HUMAN$", "\\1", prot_anno$Protein.Id)
head(prot_anno)

#' @title Extract differentially expressed proteins
#'
#' @description A wrapper around limma::topTable to extract molecules with sign.
#' changes for a certain comparison which is selected via \code{coefficient}.
#'
#' @param limma_fit2 The result of limma::eBayes
#' @param coefficient which contrast to extract, can be the name of a contrast
#' as defined in `\code{colnames(contrast.matrix)} or an integer indicating 
#' the column of \code{contrast.matrix} that should be used.
#' @param p_val_threshold set the maxmimum p-value that you will be willing to 
#' accept for determining significantly changed molecules
#' @param p_val_adj_method Specify the type of adjustment for multiple hypothesis
#' testing. Default: "BH"; other options include: "BY", "holm","hochberg", "hommel", "bonferroni"
#' @param annotation_dt a data.table with additional identifiers for the proteins;
#' should have a column named "Uniprot" so that it can be merged with the limma
#' output
#' @examples \dontrun{
#' 
#' }
extract_DP <- function(limma_fit2, coefficient, p_val_threshold, p_val_adj_method = "BH", annotation_dt){
  
  ## determine whether the annotation_dt can be used
  add_anno <- TRUE
  if(!"Uniprot" %in% names(annotation_dt)){
    warning("The object supplied via annotation_dt does not contain a column named
            'Uniprot'. This would be needed to add the information to the limma 
            output; since it is missing, the original limma output will be returned
            as a data.table.")
    add_anno <- FALSE
  }
  
  ## extract the limma results and turn them into a data.table
  results <- limma::topTable(limma_fit2, coef = coefficient, 
                             adjust = p_val_adj_method, p.value = p_val_threshold,
                             number=Inf)
  results <- as.data.table(results, keep.rownames = TRUE)
  setnames(results, "rn", "Uniprot")
  
  ## add annotation if applicable
  if(isTRUE(add_anno)){
    results <- annotation_dt[results, on = "Uniprot"]
  }
  
  return(results)
}
```

```{r}
## extracting DN vs EV
dp_res_DN_vs_EV <- extract_DP(fit2, coefficient = 2, p_val_threshold = 0.001, p_val_adj_method = "BH", annotation_dt = prot_anno)

## extracting TGFB vs EV
dp_res_TGFB_vs_EV <- extract_DP(fit2, coefficient = 3, p_val_threshold = 0.001, p_val_adj_method = "BH", annotation_dt = prot_anno)
```

```{r write_to_xlsx, eval=FALSE}
## save XLS worksheets
library(openxlsx)
wb <- createWorkbook()
addWorksheet(wb, "EV_vs_DN")
addWorksheet(wb, "EV_vs_TGFB")

writeData(wb, "EV_vs_DN", dp_res_DN_vs_EV,rowNames=FALSE)
writeData(wb, "EV_vs_TGFB", dp_res_TGFB_vs_EV, rowNames=FALSE)
```

#### Comparing DN vs TGFB DP

```{r comparison_venn, fig.width=5, fig.height=5}
## Venn diagram
#decideTests(fit2, p.value = 0.001) %>% limma::vennDiagram()

decideTests(fit2, p.value = 0.001)[,2:3]  %>% limma::vennDiagram()
```

* There are 799 (29 + 770) DP shared between TGFB and DN (= intersect).
* 2405 are changed in either one: 770 + 29 + 987 + 241 + 96 + 282 (= union of TGFB-EV and DN-EV).

```{r}
## extracting the DP shared between DN-EV and TGFB-EV
DP_intersect <- merge(dp_res_DN_vs_EV, dp_res_TGFB_vs_EV, 
                      by=c("protein", "Protein.Id", "Uniprot","gene_symbol"), 
                      all=FALSE) # previously named output: "merged_allFalse"
nrow(DP_intersect)
## replace .x with "DN" and .y with "TGFB" -- this relies on the order of the merge
## command!
setnames(DP_intersect, grep("\\.x$", names(DP_intersect), value = TRUE), gsub("x$", "DN", grep("\\.x$", names(DP_intersect), value = TRUE)))
setnames(DP_intersect, grep("\\.y$", names(DP_intersect), value = TRUE), gsub("y$", "TGFB", grep("\\.y$", names(DP_intersect), value = TRUE)))

## extracting the **universe** of DP that are changed in either comparison
DP_union <- merge(dp_res_DN_vs_EV, dp_res_TGFB_vs_EV,
                  by=c("protein", "Protein.Id", "Uniprot","gene_symbol"),
                  all=TRUE) # previously named output: merged
nrow(DP_union)
setnames(DP_union, grep("\\.x$", names(DP_union), value = TRUE), gsub("x$", "DN", grep("\\.x$", names(DP_union), value = TRUE)))
setnames(DP_union, grep("\\.y$", names(DP_union), value = TRUE), gsub("y$", "TGFB", grep("\\.y$", names(DP_union), value = TRUE)))

```


```{r write_to_xlsx2, eval=FALSE}
addWorksheet(wb, "intersect_TGFB_and_DN")
addWorksheet(wb, "union_TGFB_and_DN")
writeData(wb, "intersect_TGFB_and_DN", DP_intersect, rowNames=FALSE)
writeData(wb, "union_TGFB_DN", DP_union, rowNames=FALSE)
saveWorkbook(wb, "~/Dropbox/tetERK ABC/Px/Paola 3x Data/Total/Analysis/totalPx_exp2_p001.xlsx", overwrite=T)
```


```{r}
## combi.dt contains the original, normalized values that were used for the limma
## comparisons
##
## to add protein and gene names to the combi.dt:
setnames(combi.dt, "uniprot", "Uniprot")
combi.dt <- prot_anno[combi.dt, on = "Uniprot"]
```


## DN vs. EV
 
```{r}
## extract all the proteins with log2FC and stats values
dp_all_DN <- extract_DP(fit2, coefficient = 2,
                     p_val_threshold = 1, p_val_adj_method = "BH", annotation_dt = prot_anno)

## add information to the limma object to make use of limma's plots
fit2$genes$ID <- rownames(fit2) ## uniprot IDs
fit2$genes$Status <- ifelse(fit2$genes$ID %in% dp_res_DN_vs_EV$Uniprot, "sig.diff.inDN", "n.s.")

## if you want the protein names instead:
tmp <- data.frame(protein = prot_anno$protein, row.names= prot_anno$Uniprot)
fit2$genes$ID <- tmp[rownames(fit2),]
```




## Heatmap

For the heatmap, we will need the original values per samples (in their normalized
form). That data is kept in `combi.dt`.

```{r hm_prep}
## first, we need to get the data into a wide format
hm.mat <- dcast(combi.dt,  gene_symbol + Uniprot ~ sample, value.var = "quot_norm") %>% as.data.frame
row.names(hm.mat) <- paste(hm.mat$gene_symbol, hm.mat$Uniprot, sep = "_")
#head(hm.mat)
#dim(hm.mat)

## Heatmaps with **all** proteins
#pheatmap::pheatmap(hm.mat[, -c(1:2)], scale = "row", show_rownames = FALSE)
# turning off the clustering for the samples:
#pheatmap::pheatmap(hm.mat[, -c(1:2)], scale = "row", cluster_cols = FALSE, show_rownames = FALSE)
```

If you want to generate a heatmap for just a subset of genes, those can be chosen
via the `subset` command.

```{r hm_intersect_TGFB_DN, fig.height = 10, fig.width=5}
## subsetting the matrix to proteins of interest, e.g. those that are part of
## the intersection of TGFB and DN
hm.mat_DPintersect <- subset(hm.mat, Uniprot %in% DP_intersect$Uniprot)
nrow(hm.mat_DPintersect)
hm.mat_DPintersect$gene_symbol <- NULL
hm.mat_DPintersect$Uniprot <- NULL

## heatmap plotting & clustering
hm_clust <- pheatmap::pheatmap(hm.mat_DPintersect, scale = "row", color = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdBu")))(100), fontsize_row = 1, cutree_rows = 3, show_rownames = F)
```

```{r hm_extract_clusters, eval=FALSE}
## extract cluster IDs
## the number here should be the same as for cutree_rows in the pheatmap command
clusters <- cutree(hm_clust$tree_row, 3)

data.frame(cluster = clusters, 
           gene_symbol = gsub("_.*","", names(clusters)),
           uniprot = gsub(".*_","",names(clusters))) %>% 
  write.table(file = "DP_clusters_exp2_p001_cut3v2.txt", sep = "\t", quote = F)
```

# Over-representation analyses

For over-representation analyses, we’re going to compare the following groups (“clusters”):

genes whose proteins show changes in both TGFB and DN
up
down
clusterCompare needs a list of entrez IDs for each group.


```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)
columns(org.Hs.eg.db)
```

```{r}
anno_entrez <- select(org.Hs.eg.db, keys = unique(dp_all_DN$Uniprot), 
                      columns = c("ENTREZID", "SYMBOL"), keytype = "UNIPROT") %>% as.data.table
setnames(anno_entrez, c("UNIPROT"), c("Uniprot"))

## add ENTREZ entries to the data.tables with the shared DP
DP_intersect <- anno_entrez[DP_intersect, on = "Uniprot"]
```


```{r}
clstcomp.list <- list(up = unique(DP_intersect[logFC.DN > 0 & logFC.TGFB > 0]$ENTREZID),
                      down = unique(DP_intersect[logFC.DN < 0 & logFC.TGFB < 0]$ENTREZID))

univ.entrez <-  anno_entrez[Uniprot %in% unique(prots$Uniprot)]$ENTREZID %>% unique
```


```{r}
library(clusterProfiler)
ck.GO_cc <- compareCluster(clstcomp.list, fun = "enrichGO", OrgDb = org.Hs.eg.db, ont = "CC", universe = univ.entrez)  
```


```{r, fig.width=5, fig.width=5}
dotplot(ck.GO_cc) + 
  ggtitle("Over-represented GO terms (CC)",
          subtitle="proteins changed in DN & TGFB") +
   scale_color_gradientn(colours =rev(c("darksalmon", "firebrick3", "firebrick4")))
```